Formalizing Stack Safety as a Security Property

### 2023-01-25 submission5/intro-v2.tex
Checkpoint on intro

\rb{\textbf{The call stack and its security.} Maybe a few opening words to set
  the stage and bring two things to the front: (1) functions and subroutines as
  fundamental building blocks of any kind of structured programming; and (2) the
  stack as the natural abstraction to manage these in a computer. And also maybe
  a little bit more about the width, the gravity and the \emph{foundational
  nature} of attacks on the stack, which the sentence below touches on
  summarily.
\rb{(+SNA) Connection to functions (and adversarial code?). Attacks on the stack
  are also attacks on this fundamental abstraction (functions, subroutines,
  procedures, methods, however you call them)}
}
%
The call stack is a perennial target for low-level attacks, leading to
dire consequences from leakage and corruption of private stack data
to control-flow hijacking.
%
\rb{We can mention things like e.g. the most recent CWE Top 25 Most Dangerous
    Software Weaknesses from MITRE, cf.
    \url{https://cwe.mitre.org/top25/archive/2022/2022_cwe_top25.html}, where
    e.g. \#1 (Out-of-bounds Write), \#5 (Out-of-bounds Read) and \#7 (Use After
    Free), that is a good number of the most important and prevalent
    vulnerability types are closely related to the stack. These will show up in
    other popular rankings as well.
\rb{(+SNA) List major CWEs? No matter the class of weakness considered, many
  involve a \emph{flow} going through the stack (cf. Eternal war on memory).}
}
%
To foil such attacks, a menagerie of
software and hardware protections have been proposed,
%
including stack canaries~\cite{Cowan+98},
bounds checking~\cite{NagarakatteZMZ09,NagarakatteZMZ10,DeviettiBMZ08},
split stacks~\cite{Kuznetsov+14},
shadow stacks~\cite{Dang+15,Shanbhogue+19},
capabilities~\cite{Woodruff+14,Chisnall+15,SkorstengaardLocal,SkorstengaardSTKJFP,Georges+21},
and hardware tagging~\cite{DBLP:conf/sp/RoesslerD18}.
  \ifaftersubmission\apt{Mostly from
  nick; there could be more}\bcp{Yes, going back to MIT days---we should
  include several more of these, if only to give readers the impression that
this is a well-studied mechanism (so formalizing its protections is
useful).}
\fi
%
\rb{Did we want to add even more techniques? Is this selection broad enough?
  Anyway reference that we'll be looking at these more closely later.}

The protections offered by such mechanisms are commonly described by giving
concrete examples of attacks that they can prevent---corruption of return
addresses, buffer overflows, use of uninitialized variables, etc.---leaving
a more formal characterization to the reader's intuition.
%
\rb{Begins to anticipate subsequent discussion. This is even something of an
  understatement: some papers (esp. newer ones?), make concrete claims, though
  they remain informal. Arguably, none have a truly formal characterization in
  mind (with the qualified exception of \cite{SkorstengaardSTKJFP}): this is one
  of the novelties of this work!
\rb{(+SNA) Indeed, this point needs to be made carefully. We claim to go further
  than previous work, without showing a fully general framework (which would be
  highly parametric, etc.). Tie to the idea of micro-policies as a good place to
  start because of their flexibility, elegance, enabling a wide of interesting
  experiments, etc.}
}
%
Previous attempts
to produce a formal characterization do so using a {\it fully abstract overlay semantics},
a style of correct-by-construction machine \cite{SkorstengaardSTKJFP}.
But this approach is not suitable for the {\em lazy} enforcement mechanism
of Roessler and DeHon, in which stores are not checked, and illegal stores are
only caught when their location is later read. We aim to provide a stack-safety
property suitable for testing lazy enforcement.
%
\rb{Need to discuss overlay semantics, but also note that it is very intensional
  and very specific, as the authors acknowledge. I'd introduce this a bit more
  gradually. Also, if phrased this way, it sounds like its main drawback is a
  lack of generality, to which we should contrast a more general approach, not
  just one that is specialized to a specific policy, i.e. lazy micro-policies
  (I'd try to present these more gradually as well).}

@@

Security mechanisms designed to prevent a certain attack can be brittle---successfully
eliminating the attack as it exists at the time of publication, but leaving room for attackers
to find new, similar attacks. To mitigate this risk, mechanisms should aim to enforce formal
properties of correct behavior that can be proven or tested. Such properties become the
specification against which enforcement can be validated; even enforcement mechanisms that do
not fulfill the property benefit from the ability to articulate {\it why} they fail.
%
\rb{Good stuff! There is some overlap between this discussion and the opening
  motivation, maybe we could restructure things a bit so that those threads are
  unified?}

Stack safety is challenging to specify in full generality, because the stack is fundamentally
an architecture-specific construct. (Contrast Acevedo de Amorim et al.'s \cite{} work on heap safety,
which is broadly consistent across architectures and language semantics, and therefore amenable
to a very general approach.)
The desired behavior of the stack varies based on ISA, calling convention,
and the features of whichever programming language originated the code.
For instance, the current state-of-the-art stack-safety specification from Georges et al.
\cite{Georges22:TempsDesCerises} supposes a machine with only straightforward call-and-return
control flow, and therefore defines {\it Well-bracketed Control Flow} (WBCF) under the assumption that
a callee should always return to its caller. But, in the presence of tail-call elimination, a
return from a callee to a distant ancestor is perfectly reasonable. Any addition of new features
must be accompanied by tweaks to the model.
%
\rb{This point is extremely important!}

>>>>

\subsection{The Call Stack and Its Security}

Computer programming at any scale relies on \emph{functions} (variously known as
subroutines, procedures, methods, etc.) as a \emph{foundational abstraction} to
structure programs into building blocks representing units of computation that
can \emph{call} one another to create larger computations.
%
At a low level, a function call comprises its own data (private or shared), as
well as information about its \emph{caller} and \emph{callee} functions. The
runtime must keep track of the local contexts of all active function activations
at any given time; the \emph{fundamental data structure} charged with realizing
this abstraction is the \emph{(call) stack}.
%
\rb{This should be well known, so no need for illustration?}

The correct execution on a program---and its security---is contingent on all of
its functions properly implementing this abstraction by means of a careful use
of the call stack. Be it accidental or intentional, incorrect management of the
stack is at the root of a myriad of faults whose effects range from (rarely)
innocuous to (usually) catastrophic.
%
From a security perspective, attacks on the stack (say, from a malicious
function that chooses to abuse its access to the contents of this data
structure) are as fundamental, varied and pervasive as the functional
abstraction itself.
%
\rb{Begins to prefigure the idea of multiple stack implementations.}

%% \rb{\textbf{The call stack and its security.} Maybe a few opening words to set
%%   the stage and bring two things to the front: (1) functions and subroutines as
%%   fundamental building blocks of any kind of structured programming; and (2) the
%%   stack as the natural abstraction to manage these in a computer. And also maybe
%%   a little bit more about the width, the gravity and the \emph{foundational
%%   nature} of attacks on the stack, which the sentence below touches on
%%   summarily.
%% \rb{(+SNA) Connection to functions (and adversarial code?). Attacks on the stack
%%   are also attacks on this fundamental abstraction (functions, subroutines,
%%   procedures, methods, however you call them)}
%% }

Indeed, the call stack is a perennial target for low-level attacks, leading to
dire consequences, from leakage and corruption of private stack data
to control-flow hijacking.
%
Even a cursory glance at the most recent release of the authoritative CWE Top 25
Most Dangerous Software Weaknesses
list\footnote{\url{https://cwe.mitre.org/top25/archive/2022/2022_cwe_top25.html}}
shows general classes of vulnerabilities closely related to the stack \rb{(though not exclusively)}
consistently at and near the top, e.g., \#1 (CWE-787, Out-of-bounds Write), \#5 (CWE-125, Out-of-bounds Read), \#7 (CWE-416, Use
After Free) or \#11 (CWE-476, NULL Pointer Dereference).
%
Moreover, many other types of faults involve some kind of control or data flow
using data stored in the stack \rb{cf. Eternal war in memory, Google/Microsoft
memory safety statistics, etc.}.

To foil such attacks, a menagerie of
software and hardware protections have been proposed,
%
including stack canaries~\cite{Cowan+98},
bounds checking~\cite{NagarakatteZMZ09,NagarakatteZMZ10,DeviettiBMZ08},
split stacks~\cite{Kuznetsov+14},
shadow stacks~\cite{Dang+15,Shanbhogue+19},
capabilities~\cite{Woodruff+14,Chisnall+15,SkorstengaardLocal,SkorstengaardSTKJFP,Georges+21},
and hardware tagging~\cite{DBLP:conf/sp/RoesslerD18}---we discuss these in more
detail in \cref{XXX}.
  \ifaftersubmission\apt{Mostly from
  nick; there could be more}\bcp{Yes, going back to MIT days---we should
  include several more of these, if only to give readers the impression that
this is a well-studied mechanism (so formalizing its protections is
useful).}
\fi
%
\rb{Broad enough? Do we want to add more techniques?}
%
Yet despite this wealth of potential solutions, the stack remains a central
security concern---why?
%
Security mechanisms designed to prevent a certain attack can be brittle---successfully
eliminating the attack as it exists at the time of publication, but leaving room for attackers
to find new, similar attacks.
%
%% \rb{And so the cat-and-mouse game continues.}

To mitigate this risk, \rb{we claim that} mechanisms should aim to enforce \emph{formal
properties} of correct \rb{safe?} behavior that can be proven or tested. Such properties become the
\emph{specification} against which enforcement can be validated; even enforcement mechanisms that do
not fulfill a property benefit from the ability to articulate \emph{why} they fail.
%
%% \rb{Good stuff! There is some overlap between this discussion and the opening
%%   motivation, maybe we could restructure things a bit so that those threads are
%%   unified?}
%
This is the problem we begin to address in this paper \rb{or some such
  connecting words}.

\subsection{Security Properties for Stack Safety}

The protections offered by stack protection mechanisms are commonly described by giving
concrete examples of attacks that they can prevent---corruption of return
addresses, buffer overflows, use of uninitialized variables, etc.---leaving
a more formal characterization to the reader's intuition.
%
\rb{Begins to anticipate subsequent discussion. This is even something of an
  understatement: some papers (esp. newer ones?), make concrete claims, though
  they remain informal. Arguably, none have a truly formal characterization in
  mind (with the qualified exception of \cite{SkorstengaardSTKJFP}): this is one
  of the novelties of this work!
\rb{(+SNA) Indeed, this point needs to be made carefully. We claim to go further
  than previous work, without showing a fully general framework (which would be
  highly parametric, etc.). Tie to the idea of micro-policies as a good place to
  start because of their flexibility, elegance, enabling a wide of interesting
  experiments, etc.}
}
%
Previous attempts
to produce a formal characterization do so using a {\it fully abstract overlay semantics},
a style of correct-by-construction machine \cite{SkorstengaardSTKJFP}.
But this approach is not suitable for the {\em lazy} enforcement mechanism
of Roessler and DeHon, in which stores are not checked, and illegal stores are
only caught when their location is later read. We aim to provide a stack-safety
property suitable for testing lazy enforcement.
%
\rb{Need to discuss overlay semantics, but also note that it is very intensional
  and very specific, as the authors acknowledge. I'd introduce this a bit more
  gradually. Also, if phrased this way, it sounds like its main drawback is a
  lack of generality, to which we should contrast a more general approach, not
  just one that is specialized to a specific policy, i.e. lazy micro-policies
  (I'd try to present these more gradually as well).}

@@

Stack safety is challenging to specify in full generality, because the stack is fundamentally
an architecture-specific construct. (Contrast Azevedo de Amorim et al.'s \cite{DBLP:conf/post/AmorimHP18} work on heap safety,
which is broadly consistent across architectures and language semantics, and therefore amenable
to a very general approach.)
The desired behavior of the stack varies based on ISA, calling convention,
and the features of whichever programming language originated the code.
For instance, the current state-of-the-art stack-safety specification from Georges et al.
\cite{Georges22:TempsDesCerises} supposes a machine with only straightforward call-and-return
control flow, and therefore defines {\it Well-bracketed Control Flow} (WBCF) under the assumption that
a callee should always return to its caller. But, in the presence of tail-call elimination, a
return from a callee to a distant ancestor is perfectly reasonable. Any addition of new features
must be accompanied by tweaks to the model.
%
\rb{This point is extremely important!}
